---
title: "Intro to R"
author: "Kyla McConnell"
output: html_document
---

# Welcome to R!

Why should you learn R? 

It's...

- totally free!

- open-source

- a scripting language (reproducible and transparent)

- an established tool with tons of tutorials and help pages 

- an amazing and inclusive community 



## Step One: Install R and R-Studio

R = programming language that can be written directly into your computer's console/terminal

R-Studio = integrated development environment (IDE) with lots of features

Almost all R-users have both installed 


## Step Two: Check out R-Studio

Four panels: 

Top-left: scripting panel -- write and save code (may only open when you open a new file)

Bottom-left: console panel -- input/output that won't be saved

Top-right: environment panel -- dataframes, variables (and sometimes has other features too, depending on what you're working with)

Bottom-right: install and update packages, preview plots, read help files, and some other features


## Step Three: Open your first script

Saving files: continue working on code, re-run it later, or have as a record 

R script vs. R-Markdown

### R script (.R)

File -> New File -> R Script 

- simple: only code

- write lines of code in the order they are meant to be run
- run each line by clicking that line and pressing Control + R (on Windows) or Cmd + Enter (on Mac)
- output shows below, in console panel (number in bracket shows amount of output if there is more than one line)

- Try it out: 5 + 5 

- comments that are not read as code: #


### R-Markdown file (.Rmd)

File -> New File -> R Markdown

- R-Markdown: more complex, both code and text, can also include images and some basic formatting (## for headers)

- YAML information at top:

---
title: "Intro to R"
author: "Kyla McConnell"
date: "12/21/2020"
output: html_document
---


### How to use R-Markdown

- text in whitespace = text (unlike in an R script)
- can use text to keep notes or leave a record of your decisions

- code needs a code chunk -- type it with three backticks or use Insert in the top bar and select R
  
```{r}

```

Run code: 
- highlight and press Control + R (on Windows) or Cmd + Enter (on Mac)
- use green right-pointing arrow at right edge of code block
- arrow with line under it -> runs all code chunks above (but not the current one)

Output: 
- directly below code block in preview window

### Try it out!

1. Open a new R-Markdown file 

2. Delete or adapt the template that is provided in the file

3. Add a new (R) code block and try some basic arithmetic. What happens if you have multiple lines of code in your code block? What happens if you chain together multiple operations? Try using different types of arithmetic symbols (+, - , * , / , ^)

Bonus: Can you figure out what the %% operator does (ex: 11 %% 2, 11 %% 3)



### Variables

Variables: give name/label to output, saves it (can be used anywhere below in current script)

name_of_variable <- code

For example:
```{r}
fav_number <- 12

fav_number * 3
```



```{r}
worst_number <- 7

difference <- fav_number - worst_number
```



```{r}
difference
```



```{r}
double_number <- fav_number * 2
double_number
```


### Try it out

1. You are planning a pizza party and want to figure out how many pieces of pizza each of your guests can eat before the pizzas run out. Create a variable called "guests" and assign it to the number of guests you have at your party: 15. 
2. You order 5 pizzas. Assign the variable "pizzas" to the amount of pizzas you have: 6. Each pizza has 12 slices. Assign the variable "slices" to the amount of slices you have over all pizzas (hint: multiple the slices per pizza by the total number of pizzas -- you can use the pizzas variable here instead of retyping the number)
4. Divide your "slices" variable by your "guests" variable to figure out how many slices each person can have. 
5. 5 more people arrive to the party uninvited. Update your "guests" variable to the total number you have now. How many slices can each person have now?


### Data types

- Data types -- how R interprets the input you give it
- Figure out what type something is with `class()` 

```{r}
class(8.8)

class(double_number)
```

- Numbers generally treated as *numeric* -> can include decimals 

- *integer* allows only whole numbers -> can force numbers into this form with `as.integer()`
  - this will truncate the decimal (cut it off), not round

```{r}
as.integer(8.8)

integer_variable <- as.integer(8.8)
integer_variable

class(integer_variable)
```

- *character* is text, always enclosed in either double or single quotation marks 

```{r}
class("hello")
```

Check out the error here -- it may sound confusing but try to identify what it is telling you.

```{r}
my_word <- "hello"
class(my_word)

my_word + 2
```

- *logical* can be either TRUE or FALSE, nothing else. 

test if something is identical with ==
test if something is not identical with !=
test if something is greater than with > or less than with <
or try greater than or equals to >= or less than or equal to with <=

```{r}
7 < 8

"dog" == "dogs"

8.8 >= 5.6

"seven" == 7

class("seven" == 7)
```

*factor* - string variables that should be treated as categories / distinct levels of a grouping -> will become more relevant when we work with datasets

### Try it out

1. Assign your name to the variable "my_name". What data type is the variable?
2. Try the command `nchar()` on your name variable. What does it do? Try also toupper().
3. What does nchar(my_name) >= 4 return? What does this mean?

Bonus: What does TRUE == 1 return? What about FALSE == 1? Try them both with 0 too. What does this tell you about how R deals with logical variables?

### Vectors

Special variable type: *vector* -- contains multiple items of the same type, enclosed in the syntax `c()`

- command `length()` returns number of items in the vector

```{r}
grocery_list <- c("apples", "bananas", "bread", "orange juice")

length(grocery_list)
```

You can also do "vector-wise" operations:
```{r}
my_numbers <- c(12, 14, 26, 57, 82)

my_numbers + 2
```

Mathematical operation is "broadcast" to each item


Add/subtract/multiply/divide a vector by another vector: items matched up one at a time in a loop

```{r}
my_numbers + my_numbers
```


You can also take a certain item from a vector using square brackets. For example, take the first item:

```{r}
my_numbers[1]
```

### Try it out

1. Make a variable called long_array and assign it the values 1, 1, 2, 2, 5, 10 (don't forget the `c()` syntax!)
2. Make a variable called short_array and assign it the values 1, 6, 10
3. Subtract short_array from long_array
4. Subtract long_array from short_array
5. Look at the output. What does this tell you about how R deals with vectors of different lengths?

### Installing packages

*Packages* - extend basic R functions to add more complicated/specialized functionality

- CRAN: R Consortium's host for packages -- any advanced R-programmer can make and submit their packages, any R-user can download and use them

- Install packages once when you first use them
  - command in console (or Markdown): `install.packages("packagename")`
  - or use the panel at the lower right, under the Packages tab

Try it now with one of the most used packages:
```{r}
install.packages("tidyverse")
```

- CRAN "mirror": streamline your download by sending it from a close source

### Using packages

- `library()` command at beginning of script to use calls from that package -- will only work if you've already installed the package

- Good style: load all packages in first code block

- Error about not finding a package? You probably don't have it installed. 

```{r}
library(tidyverse)
```

### Reading in dataframes / tibbles

- What type of file are you working with?
  - Excel: go to File -> Save As and select either comma-separated values (csv) or tab-separated values (tsv)

- Where is the file saved?

  - working diretory: the location that R should start looking for files -- `setwd()`
    - ex: `setwd(~Users/kyla/Documents...)` or `setwd(C:/Documents...)`
    - can also use Session -> Set Working Directory -> Choose Directory...
  - R-Markdown automatically sets the working directory to the folder you're in
  
  - once you've set the working directoy:
    - can use files with just their short names ("experiment_data.csv")
    - can use data in a folder ("data/experiment_data.csv")
    - can use ../ to go backwards one folder

- Read in the file:
  - base-R: `read.csv()` and `read.tsv()`
  - tidyverse (improvements, reads as tibble instead of dataframe): `read_csv()` and `read_tsv()`
    - must have tidyverse installed and loaded with a `library()` call
    - note: `read_csv2()` for csv files in languages that use a comma as a decimal separator and thus a semicolon as a csv separator 
    
*Save the output to a variable.*

```{r}
jump_data <- read_csv("../data/jump.csv")
```



### Exploring your data

Now you have a data file read in, but how do you see what's in it?

`head()`: shows first six rows 
```{r}
head(jump_data)
```

Can change the amount of rows with the n *argument* 
```{r}
head(jump_data, n=3)
```

Or: click name of dataframe in Environment tab
  - can also sort columns and Filter rows -- just for viewing purposes
  - bit slow if you start having huge dataframes but often a good first look

`summary()`: call it on a dataframe to get each column and useful info based on the data type. For example, numeric columns will show the min, median, max and the quartiles (25% increments).
```{r}
summary(jump_data)
```

### Try it out

1. Load in the dataframe "spr_example.csv" as a tibble, i.e. using the tidyverse call `read_csv()` Make sure to save it to a variable.
2. Take a look at your dataframe in the Environment tab. 
3. Try the command `colnames()` on the variable you saved the dataframe to. What does it do?
4. We already tried the `head()` command, now try the `tail()` command. What does it do?

### Working with columns

- specific column call with dataframe_name$column_name

R-Studio will helpfully suggest the column names as soon as it sees the $
```{r}
jump_data$participant
```

```{r}
jump_data$height_cm
```

### Descriptive stats

Basic descriptive statistics for numeric columns -- mean, median and range 

```{r}
mean(jump_data$height_cm)
```

```{r}
range(jump_data$height_cm)
```

```{r}
median(jump_data$height_cm)
```

```{r}
summary(jump_data$height_cm)
```


### Try it out

1. Use the variable you saved the spr_example data to. What is the range of response times (RT)?
2. Save the mean RT to a variable called "average_RT". This number is in milliseconds. What would the equivalent be in seconds (hint: divide by 1000)
3. How many unique words are in this experiment? How many conditions are there?


### Categorical columns

Tibbles:
- text is defaulted to character
- factors represent groups (participant, conditions, etc.)

Summary on a character column shows that each entry is seen as different and uncomparable:
```{r}
spr_data <- read_csv("data/spr_example.csv")

class(spr_data$participant)

summary(spr_data$participant)
```

- Change a character column into a factor using `as.factor()` 
  - save over the original column

- As character, R see that there are 22 instances of each of the four labeled groups:
```{r}
spr_data$participant <- as.factor(spr_data$participant)

class(spr_data$participant)

summary(spr_data$participant)
```

We'll revisit the idea of factors later!

- Quick reference: check the data tab upper-right

### Conclusion

- Installed R, R-Studio
- R-Markdown 
- variables
- data types: numeric, integer, character, factor, vector
- installing & loading packages
- reading in files 
- looking at dataframes in R

Next time:
- tidyverse
- rename columns
- make new columns
- reorder 
- filter based on condition
- etc