---
title: "Intro to R"
author: "Kyla McConnell"
output: html_document
---

### Reading in dataframes / tibbles

- Need to know: file type, file location on computer

- Working directory 
  - `setwd()`
  - Session -> Set Working Directory -> Choose Directory...
  - R-Markdown automatically sets the working directory to the folder you're in
  
Print current working directory with `getwd()`
```{r}
getwd()
```
  
  - once you've set the working directory:
    - can use files with just their short names ("experiment_data.csv")
    - can use data in a folder ("data/experiment_data.csv")
    - can use ../ to go backwards one folder

- Read in the file:
  - tidyverse: `read_csv()` and `read_tsv()`
    - must have tidyverse installed and loaded with a `library()` call
    - `read_csv2()` for csv files in languages that use a comma as a decimal separator and thus a semicolon as a csv separator 
    
*Save the output to a variable.*

#### Try it out

1. Find the dataframes "penguins_data_1", "penguins_data_2" and "penguins_data_3" on your computer's file system. Open the files and take a look. Make note of the file extension and open the files (preferably in a text editor, not Excel) and take a look at the separator characters. Then, load all three dataframes into this R session using the appropriate calls (be sure to load tidyverse first!)

```{r}
library(tidyverse)

penguins_1 <-
penguins_2 <-
penguins_3 <-
```

2. Take a look at each of the dataframes in the Environment panel -- note how it tells you how many rows and columns each dataframe has, and what the type of the data in each column is. Did you read in the files correctly? If not, click on the name of the dataframe in the Environment panel to view the dataframe. Can you see what went wrong?

3. Try calling `head()`, `tail()` and `summary()` on one of the penguin dataframes (they are actually all the same) and see the output.


### Working with columns

- From here on, lets use just one of the penguin dataframes.

- Call just one column with: dataframe_name$column_name

```{r}
penguins <- read_csv("../data/penguin_data_1.csv")

penguins$species
```

```{r}
penguins$body_mass_g
```

### Descriptive stats

Basic descriptive statistics for numeric columns -- mean, median and range 

```{r}
mean(penguins$body_mass_g)

mean(penguins$flipper_length_mm)
```

```{r}
range(penguins$body_mass_g)

range(penguins$bill_length_mm)
```

```{r}
median(penguins$body_mass_g)
```

```{r}
summary(penguins$bill_length_mm)

summary(penguins$body_mass_g)
```


#### Try it out

1. Read in the data in "spr_example.csv" and save it as a variable
2. What is the range of response times (RT)?
3. Save the mean RT to a variable called "average_RT". This number is in milliseconds. What would the equivalent be in seconds (hint: divide by 1000)


### Categorical columns

Tibbles:
- text is defaulted to character
- factors represent groups (participant, conditions, etc.)

Summary on a character column shows that each entry is seen as different and uncomparable:
```{r}
spr_data <- read_csv("../data/spr_example.csv")

class(spr_data$participant)

summary(spr_data$participant)
```

- Change a character column into a factor using `as.factor()` 
  - save over the original column

- As character, R see that there are 22 instances of each of the four labeled groups:
```{r}
spr_data$participant <- as.factor(spr_data$participant)

class(spr_data$participant)

summary(spr_data$participant)
```



- Quick reference: check the data tab upper-right
